---
title: "Final Project cont."
subtitle: "Category Plots"
author: "Sophie Ennis"
date: "`r Sys.Date()`"
output: 
  md_document:
    toc: true
  pdf_document:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Gun Deaths in the United States cont.

## Load necessary libraries

```{r load-libraries, message=FALSE}
library(tidyverse)
library(rcis)
```

## Read the cleaned CSV

```{r read_csv}
clean_gun_deaths <- read.csv("clean_gun_deaths.csv")
head(clean_gun_deaths)
```
Read the CSV we wrote in the `css_final_project.Rmd` file.

## Plot One

```{r plot_one, cache=TRUE}
category_by_year <- clean_gun_deaths %>% 
  filter(!is.na(category)) %>% 
  group_by(year, category) %>% 
  summarize(count = n(), .groups = "drop")

ggplot(category_by_year,
       mapping = aes(
         x = factor(year),
         y = count,
         color = category,
         group = category)) +
  geom_line() +
  labs(
    title = "Gun Deaths Over Time by Age Group & Intent",
    x = "Year",
    y = "Number of Deaths",
    color = "Age & Intent Group")
```
I wanted to plot the category variable I created, which documents form of intent by age group, so I plotted how the number of deaths per category has changed over the three years documented in the dataset. This line graph reveals that three types of firearm deaths occurred the most frequently: Adult Suicide, with an average of about 12,500 deaths, Adult Homicide, with an average of about 10,000 deaths, and Elderly Suicide, with an average of about 7,000 deaths. Unfortunately, because each of these categories had such high numbers of deaths, the rest of the categories' data is somewhat unreadable. We'll address this problem in the next few plots, but for now, we'll analyze this plot. Elderly Suicide appears to be the only age group and intent category that increases throughout the timeline of the dataset. Adult Homicide and Youth Homicide appear to decrease throughout the timeline, while the other age group and intent categories appear to remain constant, but it is hard to tell with the bottom few lines because so many of them are on top of each other. 

```{r table, cache=TRUE}
category_summary <- category_by_year %>% 
  group_by(category) %>% 
  summarize(
    mean = mean(count, na.rm = TRUE),
    min = min(count, na.rm = TRUE),
    max = max(count, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(
  category_summary,
  caption = "Summary Statistics of Gun Deaths by Category")
```
I examined the summary statistics of these categories in a table, calculating the mean, minimum, and maximum number of deaths per year. This table gives us a better picture as to what the average number of deaths per category actually are. We can see that I estimated the averages of the categories with the highest number of deaths somewhat accurately. This table also helps us see the average number of deaths in the categories that were less visible in the original plot. For instance, now we can see that the lowest average number of deaths include Elderly Accident and Youth Accident. We can also see that suicide was the intent that killed the highest number of adults and elderly individuals, but killed only half as many people as homicide did in the Youth categories. From the minimums and maximums, we can see that the number of deaths did not change much from year to year, with the biggest difference being an average of 6,793 and a maximum of 7,133 deaths in the Elderly Suicide category, reflecting what we saw in the original plot. 

## Plot Two

```{r plot_two, cache=TRUE}
category_by_year_low <- clean_gun_deaths %>% 
  filter(!category %in% c("Adult Suicide", "Adult Homicide", "Elderly Suicide")) %>% 
  filter(!is.na(category)) %>% 
  group_by(year, category) %>% 
  summarize(count = n(), .groups = "drop")

ggplot(category_by_year_low,
       mapping = aes(
         x = factor(year),
         y = count,
         color = category,
         group = category)) +
  geom_line() +
  labs(
    title = "Gun Deaths Over Time by Age Group & Intent Excluding Top Predictors",
    x = "Year",
    y = "Number of Deaths",
    color = "Age & Intent Group")
```
I excluded the three categories with the highest number of deaths - Adult Suicide, Adult Homicide, and Elderly Suicide - to get a better look at the less visible categories in the original plot. Here, we can see that Youth Homicide now has the highest number of deaths, with an average of about 1,000 deaths each year. Youth Homicide decreases by a substantial amount from 2012 to 2013, and then increases minimally in 2014. Youth Suicide has the second highest number of deaths, which only increases to about 750 deaths in 2014. Clearly, gun deaths affecting those under the age of 18 is a problem. Elderly Homicide, Undetermined, and Youth Accident all seem to decrease by 2014, but Adult Accident and Elderly Accident appear to increase by 2014, with similar lulls to Youth Homicide in 2013.

## Plot Three

```{r plot_three, cache=TRUE}
category_by_year_suicide <- clean_gun_deaths %>% 
  filter(category %in% c("Youth Suicide", "Adult Suicide", "Elderly Suicide")) %>% 
  group_by(year, category) %>% 
  summarize(count = n(), .groups = "drop")

ggplot(category_by_year_suicide,
       mapping = aes(
         x = factor(year),
         y = count,
         color = category,
         group = category)) +
  geom_line() +
  labs(
    title = "Gun Deaths by Suicide Over Time by Age Group",
    x = "Year",
    y = "Number of Deaths",
    color = "Age & Intent Group")
```
Per one of my research questions, I wanted to investigate if certain demographic characteristics lend themselves to a likelihood to commit suicide. In this plot, I investigate if age group may play a role in this likelihood. Indeed it does, with Adult Suicide resulting in nearly 15,000 deaths every year in this timeline. Elderly Suicide averaged about 7,000 deaths per year, while Youth Suicide averaged only about 700 deaths per year. This indicates that adults are the most likely age group to commit suicide, and youths under 18 are the least likely. Elderly Suicide, however, is the only age group where deaths by suicide are substantially increasing year to year. 

## Plot Four

```{r plot_four, cache=TRUE}
category_by_year_adult <- clean_gun_deaths %>% 
  filter(category %in% c("Adult Homicide", "Adult Suicide", "Adult Accident")) %>% 
  group_by(year, category) %>% 
  summarize(count = n(), .groups = "drop")

ggplot(category_by_year_adult,
       mapping = aes(
         x = factor(year),
         y = count,
         color = category,
         group = category)) +
  geom_line() +
  labs(
    title = "Adult Gun Deaths Over Time",
    x = "Year",
    y = "Number of Deaths",
    color = "Age & Intent Group")
```
Given this result that adults are the most likely to commit suicide, I wanted to plot the other forms of intent committed by adults and compare the number of deaths they produce. Again, we see Adult Suicide with the highest number of deaths at nearly 15,000 deaths per year. Adult Homicide follows not too far behind with about 10,000 deaths per year, both of these forms of intent decreasing at nearly the same speed. Adult Accident is one of the categories with the lowest number of deaths, at about 350 deaths, but it is just barely increasing. Therefore, we can conclude that not only are adults the most likely to commit suicide, but they are more likely to commit suicide than any other intent. 

## Plot Five

```{r plot_five, cache=TRUE}
category_by_year_adult <- clean_gun_deaths %>% 
  filter(category %in% c("Youth Homicide", "Youth Suicide", "Youth Accident")) %>% 
  group_by(year, category) %>% 
  summarize(count = n(), .groups = "drop")

ggplot(category_by_year_adult,
       mapping = aes(
         x = factor(year),
         y = count,
         color = category,
         group = category)) +
  geom_line() +
  labs(
    title = "Youth Gun Deaths Over Time",
    x = "Year",
    y = "Number of Deaths",
    color = "Age & Intent Group")
```
To contrast with this comparison of intents used by adults, I wanted to examine the number of deaths produced by different forms of intent used by youths under 18 years old. Unlike adults and the elderly, where suicide is most frequently used, Youth Homicide appear to result in the highest number of deaths among youths each year, with an average of about 1,000 deaths. This number dropped in 2013, but was again on the rise in 2014. Youth Suicide results in the second highest number of deaths among youths, with an average of about 700 deaths. This number unfortunately rose in both 2013 and 2014. Youth Accident represents the smallest number of deaths at about 90 deaths, and luckily declined from 2013 to 2014.


