---
title: "Final Project"
subtitle: "Computing for the Social Sciences"
author: "Sophie Ennis"
date: "`r Sys.Date()`"
output: 
  pdf_document:
  md_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Gun Deaths in the United States

## Research Questions

My research question asks: How do victims of gun violence differ demographically? Are some members of the population more likely to be shot (by themselves or others) than others? Are some individuals' deaths more likely to have resulted from an officer-involved shooting? Are some individuals more likely to commit suicide?

## Project Description

I will be examining the dataset `gun_deaths` to look for indication that some members of the population, based on their demographics, are more likely to die by firearm than others, and moreover, that some are more likely to die at the hands of a police officer's firearm, or even their own hands. `gun_deaths` largely documents categorical variables (intent, sex, race, place, education), making bar plots and histograms very useful (and scatterplots far less so) for displaying data from this dataset. 

## Data Description

The data is from the rcis library, sourced from [Five Thirty Eight](http://fivethirtyeight.com/features/gun-deaths/). The dataset includes records on all firearm deaths in the United States from 2012 to 2014. The variable descriptions are as follows:

- id: Row number of the observation
- year: Year the death occurred
- month: Month the death occurred
- intent: Intent of the death (homicide, suicide, accidental, unknown)
- police: Indicator if the death was the result of an officer-involved shooting
- sex: Victim's sex
- age: Victim's age
- race: Victim's race (White, Black, Hispanic, Asian/Pacific Islander, Native American/Native Alaskan)
- place: Location where the shooting occurred
- education:Highest level of education of the victim

## Load necessary libraries

```{r load-libraries, message=FALSE}
library(tidyverse)
library(rcis)
```

## Explore the data

```{r explore-data, message=FALSE}
data(gun_deaths)
head(gun_deaths)
glimpse(gun_deaths)
?gun_deaths
```

## View the data in table format

```{r table, cache=TRUE}
knitr::kable(
  gun_deaths[1:5,],
  caption = "Gun Deaths")
```
This table shows the first five cases of the `gun_deaths` dataset, all from the beginning of 2012. While we are already seeing variation in the age, place, education variables, the intent, police, sex, and race variables even in a sample of the dataset reveal larger patterns about gun deaths - that they are largely experienced by white males without police involvement. While we cannot make assumptions about patterns of intent quite yet, this table shows that all five of these victims of gun violence died by suicide. 

## Clean the data

```{r clean_police, cache=TRUE}
clean_gun_deaths <- gun_deaths %>% 
  mutate(police_recode = case_when(
    police == 1 ~ "Yes",
    police == 0 ~ "No",
    TRUE ~ NA_character_)) %>% 
  select(-police)
head(clean_gun_deaths)
```
To clean the data, I started off by recoding the binary `police` variable, a column of doubles, into a string of "Yes" or "No", a column of characters, representing police presence in firearm deaths more clearly to ensure readability for the plots. I removed the original police variable to eliminate redundancy using `select()`. 

```{r unique_intent, cache=TRUE}
unique(clean_gun_deaths$intent)
```
For the next step of cleaning, I needed to determine what the unique observations were within the intent column so I could create a new column grouped by intent and age group.

```{r clean_category, cache = TRUE}
clean_gun_deaths$category <- NA

for (i in 1:nrow(clean_gun_deaths)) {
  age <- clean_gun_deaths$age[i]
  intent <- clean_gun_deaths$intent[i]
  
  if (!is.na(age) && !is.na(intent)) {
    if (age <= 18 && intent == "Suicide") {
      clean_gun_deaths$category[i] <- "Youth Suicide"}
    else if (age > 18 && age < 60 && intent == "Suicide") {
      clean_gun_deaths$category[i] <- "Adult Suicide"}
    else if (age >= 60 && intent == "Suicide") {
      clean_gun_deaths$category[i] <- "Elderly Suicide"}
    else if (age <= 18 && intent == "Accidental") {
      clean_gun_deaths$category[i] <- "Youth Accident"}
    else if (age > 18 && age < 60 && intent == "Accidental") {
      clean_gun_deaths$category[i] <- "Adult Accident"}
    else if (age >= 60 && intent == "Accidental") {
      clean_gun_deaths$category[i] <- "Eldery Accident"}
    else if (age <= 18 && intent == "Homicide") {
      clean_gun_deaths$category[i] <- "Youth Homicide"}
    else if (age > 18 && age < 60 && intent == "Homicide") {
      clean_gun_deaths$category[i] <- "Adult Homicide"}
    else if (age >= 60 && intent == "Homicide") {
      clean_gun_deaths$category[i] <- "Elderly Homicide"}
    else {
      clean_gun_deaths$category[i] <- "Undetermined"}}
  else {
    clean_gun_deaths$category[i] <- NA}}

head(clean_gun_deaths)
```

I called this new column `category` which classifies each row by the age and intent of the individual victim in that row. I first initialized the new column with `NA` for all rows. I then created a for loop to iterate through each indexed row in the dataset. I extracted the age and intent variables from the dataset for the current row. I then use a conditional statement to ensure that the age and intent of each row is available, and if either are missing, the row is skipped and the category variable is set to `NA`. The nested if/else statements then classify the row into categories based on an age range and intent. For instance, if the victim is under 18 years old and the intent is suicide, the category is set to `Youth Suicide`. If none of the conditions are met, the category is set to `Undetermined`, and if the age or intent variables are missing for that row, the category is set to `NA`. I created this column to generalize findings by age group, the plots for which can be seen in the `category_plots.Rmd` file in this repository. 

## Write the CSV

```{r write_csv}
write.csv(clean_gun_deaths, "clean_gun_deaths.csv", row.names = FALSE)
```

I wrote the cleaned dataset to a CSV.

## View the clean data in table format

```{r table_clean, cache=TRUE}
knitr::kable(
  clean_gun_deaths[1:5,],
  caption = "Clean Gun Deaths")
```
While we still cannot make assumptions about the dataset from these five cases, we can more easily see that all of these suicides occurred among adults or the elderly, and none occurred among youth, as well as that none of these instances of gun violence involved police officers. 

## Plot One

```{r avg_age, cache=TRUE}
age_intent <- clean_gun_deaths %>%
  filter(!is.na(intent)) %>% 
  group_by(intent) %>%
  summarize(avg_age = mean(age, na.rm = TRUE))
age_intent

knitr::kable(
  age_intent,
  caption = "Intent by Average Age")
```
For the first plot, I wanted to analyze the relationship between the intent behind the firearm death and the average age that each form of intent was committed at. Therefore, I filtered out the missing values, grouped by the intent variable, and calculated the mean age for each intent. I then generated a table for these summary statistics and found that homicide tends to produce the youngest victims, with an average age of about 33 years old, suicide produces the oldest victims by far, with an average age of about 50 years old. 

```{r plot_one, cache=TRUE}
ggplot(age_intent %>% 
         filter(!is.na(avg_age) & !is.na(intent)),
       mapping = aes(
         x = intent,
         y = avg_age,
         fill = intent)) +
  geom_col() +
  labs(
    title = "Gun Death Intent by Average Age",
    x = "Intent",
    y = "Average Age",
    fill = "Intent")
```
I wanted to plot these results from the table I generated, and this bar plot confirms that suicide has, by at least 10 years more than the other forms of intent, the oldest victims, averaging to about 50 years old. While accidents and undetermined intent hover in the middle with accidents slightly lower, homicide continues to produce the youngest victims, averaging to just above 30 years old. 

## Plot Two

```{r female_vics, cache=TRUE}
count_female_vics <- function(data) {
  female_vics <- data %>%
    filter(sex == "F") %>%
    count()
  return(female_vics)
}
count_female_vics(clean_gun_deaths)
```
Because we only saw one female victim represented in our initial table, and based on public knowledge that males are more likely to be victims of all kinds of firearm deaths but particularly homicide and suicide, I wanted to see how many gun violence victims in the dataset were female. So, I utilized a function to filter the sex variable to only count rows that were labeled with the string `"F"` to indicate female victims. Out of the 100,798 rows and thus victims in this dataset, only 14,449 were female victims of gun violence.

```{r plot_two, cache=TRUE}
ggplot(clean_gun_deaths %>% 
         filter(!is.na(intent)),
       mapping = aes(
         x = sex,
         fill = intent)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Gun Deaths by Sex & Intent",
    x = "Sex",
    y = "Number of Deaths",
    fill = "Intent")
```
I wanted to confirm my suspicious and public knowledge that most firearm deaths involve male victims who die by suicide or homicide, so I plotted the relationship between the sex and intent variables. Suicide is clearly the most frequent form of intent for both males and females, but male victims of suicide are far more frequent than female. The same goes for the runner up, homicide, where suicide is almost twice as likely to occur than homicide within each sex, but males are much more likely to be victims of homicide than females. Accidental deaths and deaths by an undetermined intent are negligible.

## Plot Three

```{r plot_three, cache=TRUE}
ggplot(clean_gun_deaths %>% 
         filter(!is.na(education)),
       mapping = aes(
         x = sex,
         fill = education)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Gun Deaths by Sex & Education of Victim",
    x = "Sex",
    y = "Number of Deaths",
    fill = "Education Level")
```
To add another layer of complexity to the analysis of gun violence victims by sex, I wanted to analyze each sex by the group's level of education to see if education level might be an indicator as to who becomes victimized by firearms. While victims of both sexes were most likely to only have a high school degree or GED, indicating that these victims were not very educated, beyond this point is where the sexes diverge. Male victims were more likely to not have graduated high school than female victims, with this being the second largest category for males and the lowest for females. This means that male victims are less educated than female victims, pointing to poor education as an indicator for being victimized by gun violence. Next for the sexes is some college, indicating that they dropped out before completing, which is the second highest category for female victims and third highest category for male victims. The lowest education level for male victims is BA+, indicating an advanced degree, while BA+ is the third highest education level for female victims. This plot shows that female victims are more educated than male victims, which may be one factor contributing to the fact that males are more likely to be victimized by gun violence.

## Plot Four

```{r plot_four, cache=TRUE}
ggplot(clean_gun_deaths,
       mapping = aes(
         x = age,
         fill = police_recode)) +
  geom_histogram(position = "identity") +
  facet_wrap(~ race) +
  labs(
    title = "Age Distribution of Gun Deaths by Race & Police Involvement",
    x = "Age",
    y = "Number of Deaths",
    fill = "Police Involvement")
```
I wanted to analyze police involvement in gun deaths by race, so I used my recoded police variable in relationship with the age variable and faceted the plot by race. We can see that victims of gun violence are most likely to be white, with an average age of about 60 years old, with black individuals the second most likely to be victimized by gun violence but tending to be much younger, averaging at about 25 years old. Hispanics are less likely to be victims, but still peak at about 25 years old. Asians/Pacific Islanders and Native Americans are the least likely to be victims of gun violence. All of this considered, it appears based on this plot that nearly no instances of gun deaths involve police officers. In the majority of cases across all races, it appears that police are largely not involved in or responsible for gun deaths, but what happens when they are? While this plot gives us a good idea of how race plays a part in the frequency of gun deaths, it does not give us a clear picture of what happens when police are involved. So, I utilized a boxplot to represent police involvement more clearly.

## Plot Five

```{r plot_five, cache=TRUE}
ggplot(clean_gun_deaths,
       mapping = aes(
         x = race,
         y = age,
         fill = police_recode)) +
  geom_boxplot() +
  labs(
    title = "Age Distribution of Gun Deaths by Race & Police Involvement",
    x = "Race",
    y = "Age",
    fill = "Police Involvement") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
In this boxplot, we have less indication of which races are most affected by gun violence, but more indication that gun violence involving police officers affects racial identities differently. While we could barely see any evidence of police involvement in the previous plot, because this plot only looks at age rather than both age and number of deaths in relationship with the race variable, we are more able to distinguish between cases involving or not involving a police presence. For Asian/Pacific Islander, Native American, and White victims, police involvement in their deaths is much less likely to occur. For Hispanic victims, the likelihood of police being involved in their gun deaths is about the same as the likelihood of them not being involved. Only for Black victims is police involvement in their deaths more likely to occur. From this plot we can also see more clearly that White victims are more likely to die from gun violence when they are older, while Black gun violence victims not involving the police have the lowest minimum age, though they also have many much older outliers. Interestingly, Native American victims with police involvement in their deaths have the smallest and one of the lowest age ranges, with a median of about 29 years old. While Hispanic gun violence victims not involving the police have a larger interquartile range of age, the median age within this group is also about 29 years old.

## Plot Six

```{r plot_six, cache=TRUE}
race_by_place <- clean_gun_deaths %>% 
  filter(!is.na(place)) %>% 
  filter(!place %in% c("Farm", "Industrial/construction", "Other specified", 
                       "Other unspecified")) %>% 
  group_by(race, place) %>% 
  summarize(count = n(), .groups = "drop")

ggplot(race_by_place,
       mapping = aes(
         x = race,
         y = count,
         fill = place)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Distribution of Gun Deaths by Race & Location",
    x = "Race",
    y = "Number of Deaths",
    fill = "Location of Death") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1))
```
As a continued exploration into the race variable, I wanted to investigate the relationship between the race of the victims and the locations where they die. The victims' homes stands out as the most likely place for them to die for each racial identity, but taking up significantly more space for White victims than other racial identities. The locations where Asian/Pacific Islander and Native American victims die from gun violence appear to stop at the home, perhaps because the sample size for these racial groups is so small. The street is the second most likely location for Black, Hispanic, and White victims, with Black victims being the most likely to die on the street. White victims have the second largest number of deaths that occur on the street despite this location making up less of this racial group's total makeup of deaths, followed by Hispanic victims. Schools and trade/service areas follow, with Black, White, and Hispanic victims having nearly the same likelihood of dying at school, and White victims being the most likely to die in a trade/service area (such as a store) with Black victims close behind. Gun deaths in schools and trade/service areas may indicate mass casualty events, where all races are somewhat equally likely to be victimized. Looking at intent alongside these variables may clarify this.

## Plot Seven

```{r plot_seven, cache=TRUE}
intent_by_place <- clean_gun_deaths %>% 
  filter(!is.na(place) & !is.na(intent)) %>% 
  filter(!place %in% c("Farm", "Industrial/construction", "Other specified", 
                       "Other unspecified")) %>% 
  group_by(intent, place, race) %>% 
  summarize(count = n(), .groups = "drop")

ggplot(intent_by_place,
       mapping = aes(
         x = intent,
         y = count,
         fill = place)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Distribution of Gun Deaths by Intent & Location",
    x = "Intent",
    y = "Number of Deaths",
    fill = "Location of Death") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~ race)
```
Therefore, I plotted the relationship between the intent and location variables and faceted by race. White victims of gun violence are far more likely than any other racial group to commit suicide. Both White and Native American victims are most likely to commit suicide compared to other forms of intent, but the number of Native Americans (and Asians/Pacific Islanders for that matter) that have died because of a firearm is very small. Both Black and Hispanic victims of gun violence are most likely to killed in a homicide, though Black victims are the most likely to die by homicide out of any racial group. Homicide is closely followed by suicide for Hispanic victims. This plot confirms that most gun deaths for all racial groups occur in the home, with the street trailing behind for White and Hispanic victims, while Black victims are far more likely to die on the street than any other racial group. Schools and trade/service areas make a small appearance for Black, Hispanic, and White victims, and while trade/service areas are more conducive to suicides for White victims, they are more conducive to homicides for Black and Hispanic victims. While we cannot tell from this dataset if these homicides were part of mass casualty events, it appears that Black and Hispanic victims of gun violence may be more likely to die in mass shootings than White victims when looking at the public locations of where these homicides occurred. Future research may want to consider exploring the relationship between race, intent, and location in the context of mass shootings. 

## Conclusion

My research question asked if certain demographic characteristics make some members of the population more likely to die by gun violence. I also asked if demographic characteristics played a role in the likelihood of dying at one's own hands by suicide or by a police officer's hands. From my analysis, it appears that white male adults with a high school education are the most likely victims of gun violence. Suicide is the most frequent cause of death from gun violence across all races and sexes, but it is most likely to be committed by white male adults averaging around 50 years old. When police are involved in and thus responsible for a gun violence victim's death, however, the victim is most likely a Black adult, averaging around 30 years old.

## Resources

- I primarily used the provided lecture slides to aid in my exploratory data analysis.
- I used ChatGPT to debug the use of `&&` to check for logical values in the lines such as `(age <= 18 && intent == "Suicide")` in the conditional statements used to create the category variable.
- I used ChatGPT to generate the code `theme(axis.text.x = element_text(angle = 45, hjust = 1))` which sets the x axis labels to a 45 degree angle so they are more readable when the labels are longer.
- I used ChatGPT to figure out that I needed to set both the color and the group equal to the category variable in the category plots in order for the line plots to appear properly. 